Fractional Types

### 2013-02-05 frac.tex
a bit on sec3

Even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to briefly outline an interpretation of
{{langRevTF}} in \emph{biproduct dagger compact closed categories}. These
categories were introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and then by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207}. The bare definitions are
collected in Appendix~\ref{app:cat}.

The simplest example of such categories is the category of finite sets and
relations on which we focus. To interpret {{langRevTF}} in this category, we
associate a set with each type. For {{1}}, we associate a set with one
element and we interpret {{*}} as the cartesian product and {{+}} as the
disjoint union. For {{1/b}}, we associate the same set associated with {{b}}
and hence, in our denotations, we identify {{b}} and {{1/b}} as well as {{v}}
and {{1/v}}. It is intuitively clear that this category is compact closed as
any relation between $A$ and $B$ can be represented as a finite set of type
{{A * B}}. For example, consider the ``boolean negation'' relation:

{{ {[ (1/left(), right()), (1/right (), left ()) ]} }}

This is clearly isomorphic to the corresponding set of pairs. The map
{{eta_b}} constructs the identity relation at type {{b}}. For example, {{eta_{1+1} }}
is the relation:

@@

@@

\begin{definition}[Denotation of Value Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
\end{array}\]
\end{definition}

@@

A \emph{dagger category} is a category together with an involutive,
identity-on-objects, contravariant functor $\dagger$. Concretely, this means
that to every morphism $F : A \rightarrow B$, one associates a morphism
$f^{\dagger} : B \rightarrow A$, called the \emph{adjoint} of $f$, such that
for all $f : A \rightarrow B$ and $g : B \rightarrow C$, we have:
\[\begin{array}{rcl}
\mathit{id}^\dagger_A &=& \mathit{id}_A \\
(g \circ f)^\dagger &=& f^\dagger \circ g^\dagger \\
(f^dagger)^\dagger &=& f
\end{array}\]

>>>>

Even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to briefly outline an interpretation of
{{langRevTF}} in the category of finite sets and relations. This category is
an example of a \emph{biproduct dagger compact closed category} which are
categories introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} The category intuitively supports
higher-order relations as any relation between $A$ and $B$ can be represented
as a subset of {{A * B}}, i.e., as a value. More generally, the two main
features of these biproduct dagger compact closed categories are that they
support an adjoint for each morphism (to model reversibility) and a notion of
higher-order relation (built from the dagger which is modeled by our
fractional types).

To interpret {{langRevTF}} in this category, we associate a finite set
$\llbracket b \rrbracket$ with each type $b$ and a relation $\llbracket c
\rrbracket$ with each combinator $c$. The denotation of types is
straighforward: the denotation of {{1/b}} is an isomorphic copy of the
denotation of {{b}}.

\begin{definition}[Denotation of Value Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a finite set of values as follows:
\[\begin{array}{rcl}
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \} \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
\end{array}\]
\end{definition}

Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
the relation:

@@

Perhaps get rid of the semantics below and replace by a semantics which maps
each combinator to a relation. This is probably clearer than the operational
style.

@@

@@

A \emph{dagger category} is a category together with an involutive,
identity-on-objects, contravariant functor $\dagger$. Concretely, this means
that to every morphism $F : A \rightarrow B$, one associates a morphism
$f^{\dagger} : B \rightarrow A$, called the \emph{adjoint} of $f$, such that
for all $f : A \rightarrow B$ and $g : B \rightarrow C$, we have:
\[\begin{array}{rcl}
\mathit{id}^\dagger_A &=& \mathit{id}_A \\
(g \circ f)^\dagger &=& f^\dagger \circ g^\dagger \\
(f^\dagger)^\dagger &=& f
\end{array}\]
